---
title: "COVID-19 Plots"
output:
  html_document:
    toc: true
  pdf_document:
    toc: true
urlcolor: blue
bibliography: /Users/dinger/TeX/krd.bib
csl: /Users/dinger/Pandoc/bibstyles/american-geophysical-union.csl
---


```{r setup, cache=FALSE, echo=FALSE}
library(knitr)
knitr::opts_chunk$set(cache = FALSE, cache.lazy = FALSE, warning=FALSE,
                      fig.width=8.5, fig.height=6, autodep=TRUE, echo=FALSE)
```

# Intro

I liked the Financial Times plots for tracking the evolution of COVID-19
(https://www.ft.com/coronavirus-latest), but then they changed to different
plots.  So here I am more-or-less reproducing those plots (and adding some
others).  This is generated from an `Rmarkdown` document that I'll be rerendering
daily.

# Country-level data

For now using the country data from https://covid.ourworldindata.org:

```{r getdata, echo=TRUE}
cases = read.csv("https://covid.ourworldindata.org/data/ecdc/total_cases.csv",
                 stringsAsFactors=FALSE)
cases$date = as.POSIXct(cases$date)
cases$doy = round(as.numeric(difftime(cases$date, as.POSIXct("2019-12-31"), units="days")))

deaths = read.csv("https://covid.ourworldindata.org/data/ecdc/total_deaths.csv",
                  stringsAsFactors=FALSE)
deaths$date = as.POSIXct(deaths$date)
deaths$doy = round(as.numeric(difftime(deaths$date, as.POSIXct("2019-12-31"), units="days")))

# World population data from worldbank.org:
temp = tempfile()
download.file("http://api.worldbank.org/v2/en/indicator/SP.POP.TOTL?downloadformat=csv", temp)
pop = read.csv(unz(temp, "API_SP.POP.TOTL_DS2_en_csv_v2_1593924.csv"),
               skip=4, header=TRUE, stringsAsFactors=FALSE)
rownames(pop) = pop$Country.Name
#deaths.cty = colnames(deaths)
#deaths.cty = gsub(".", " ", deaths.cty, fixed=TRUE)
#ctymap = match(deaths.cty, rownames(pop))
# deaths.cty[which(is.na(ctymap))]
unlink(temp)
```

```{r functionDef}
# function that finds time when (piece-wise linear in time, log(deaths)) deaths
# exceeds a threshold D0
findTime = function(cty, deaths, D0) {
  i1 = max(which(deaths[,cty] <= D0))
  i2 = i1 + 1
  t = deaths$doy
  D = deaths[,cty]
  return(t[i1] + ( (log10(D0) - log10(D[i1])) / (log10(D[i2]) - log10(D[i1])) )*(t[i2] - t[i1]) )

}



# function that makes all plots except Naive CFR
# deaths: deaths, cases, us.deaths, or us.cases: matrix of cumulative deaths or cases
# ctys: ctys (countries) or sts (states): list of countries or states to plot
# cols: colors to use for each country or state
# dc.type: "death" or "case"
# cum.type: "Cumulative" or "Daily"
# yaxt: "log" or "lin"
# pc.type: "per.capita" or "absolute"
# pops: vector of populations, will be accessed as pops[cty]
mkPlot = function(deaths, ctys, col, dc.type, cum.type, lag=1, yaxt, day0=NULL, D0=10, xlab=NULL,
                  pc.type="absolute", pops=NULL, plotAll=FALSE) {


  adj = c(-0.1, 0.5)

  if (cum.type == "Daily") {
    deaths1 = diff(as.matrix(deaths[,ctys]), lag=lag)/lag
  } else {
    deaths1 = deaths[,ctys]
    lag = 0
  }

  if (pc.type == "per.capita") {
    deaths1 = deaths1 / matrix(pops[ctys], dim(deaths1)[1], dim(deaths1)[2], byrow=TRUE)
  }

  if (is.null(day0)) {
    day0 = sapply(ctys, findTime, deaths=deaths, D0=D0)
    xlab = paste("Days since ", D0, "th ", dc.type, sep="")
  }

  xlim = c(0, max(deaths$doy) - min(day0))
  if ("China" %in% ctys) {
    xlim = c(0, max(deaths$doy) - sort(day0)[2] + 5)
  } else {
    xlim = c(0, max(deaths$doy) - sort(day0)[1] + 5)
  }
#  ylim = c(min(sapply(ctys, function(cty, deaths){min(deaths[,cty], na.rm=TRUE)}, deaths=deaths1), na.rm=TRUE),
#           max(sapply(ctys, function(cty, deaths){max(deaths[,cty], na.rm=TRUE)}, deaths=deaths1), na.rm=TRUE))
  ylim = c(D0,
           max(sapply(ctys, function(cty, deaths){max(deaths[,cty], na.rm=TRUE)}, deaths=deaths1), na.rm=TRUE))


  ndates = length(deaths$date)

  if (dc.type == "death") {
    main = paste(cum.type, "deaths",
                 ifelse(lag > 1, paste("(", lag, "-day trailing average)", sep=""), ""),
                 "through", max(deaths$date))
    ylab = paste(cum.type, "deaths")
  } else {
    main = paste(cum.type, "cases",
                 ifelse(lag > 1, paste("(", lag, "-day trailing average)", sep=""), ""),
                 "through", max(deaths$date))
    ylab = paste(cum.type, "cases")
  }


  if (yaxt == "log") {
    linlogplot(1, 1, typ='n', xlim=xlim, ylim=ylim, xlab=xlab, ylab=ylab, main=main, grid=TRUE)
  } else {
    plot(1, 1, typ='n', xlim=xlim, ylim=ylim, xlab=xlab, ylab=ylab, main=main)
    grid(NULL, lty="solid", col=gray(0.8), lwd=0.3)
  }

  for (icty in 1:length(ctys)) {
    lines(deaths$doy[(lag+1):ndates] - day0[icty], deaths1[, ctys[icty]], col=col[icty])
    points(deaths$doy[(lag+1):ndates] - day0[icty], deaths1[, ctys[icty]], col=col[icty], cex=0.5, pch=16)
  }

  ilabel = pmin(ndates, floor(day0 + xlim[2])) - lag
  text(deaths$doy[ilabel+lag] - day0,
       as.numeric(deaths1[cbind(ilabel, match(ctys, colnames(deaths1)))]),
       ctys, col=col, adj=adj, cex=0.6)

  if (yaxt == "log" && cum.type == "Cumulative") {
    tDoubles = c(1, 2, 3, 7)
    lapply(tDoubles, function(tDouble, D0) {
                       abline(log10(D0), log10(2)/tDouble, lty="dashed", col=gray(0.5))
                     }, D0=D0)
  }

  legend("topleft", legend=ctys, col=col, pch=16, cex=0.5, lty="solid")

}


```

# Country-level deaths and cases {.tabset .tabset-fade}

```{r setupCountries}
ctys = c("United.States", "United.Kingdom", "New.Zealand",
         "France", "Germany", "Italy", "Spain", "Sweden",
         "China", "South.Korea", "Japan",
         "Iran", "India",
         "Brazil", "Mexico")

whiten = function(col, f){rgb(t((1-f)*col2rgb(col) + f*rep(255, 3)), maxColorValue=255)}
col = c(sapply(c(0.0, 0.1, 0.2)*2, function(f){whiten("red1", f=f)}),
        sapply(c(0.0, 0.1, 0.2, 0.3, 0.4)*2, function(f){whiten("blue1", f=f)}),
        sapply(c(0.0, 0.1, 0.2)*2, function(f){whiten("cyan1", f=f)}),
        sapply(c(0.0, 0.1)*2, function(f){whiten("green1", f=f)}),
        sapply(c(0.0, 0.1)*2, function(f){whiten("orange1", f=f)}))
```


## Daily deaths {.tabset .tabset-fade}

### Smoothed {.tabset .tabset-fade}

#### Log-scale

```{r plotSmoothedDailyDeaths}
# we'll use this later for the time-scale for some other plots
D0 = 10
d.day0 = sapply(ctys, findTime, deaths=deaths, D0=D0)


mkPlot(deaths, ctys, col, "death", "Daily", lag=7, yaxt="log",
       day0=d.day0, D0=1, xlab="Days since 10th death")

```

#### Linear scale

```{r plotSmoothedDailyDeathsLinear}

mkPlot(deaths, ctys, col, "death", "Daily", lag=7, yaxt="lin",
       day0=d.day0, D0=1, xlab="Days since 10th death")

```


### Raw {.tabset .tabset-fade}

#### Log-scale

```{r plotDailyDeaths}

mkPlot(deaths, ctys, col, "death", "Daily", lag=1, yaxt="log",
       day0=d.day0, D0=1, xlab="Days since 10th death")

```

#### Linear-scale

```{r plotDailyDeathsLinear}

mkPlot(deaths, ctys, col, "death", "Daily", lag=1, yaxt="lin",
       day0=d.day0, D0=1, xlab="Days since 10th death")


```


## Daily new cases {.tabset .tabset-fade}

### Smoothed {.tabset .tabset-fade}

#### Log-scale

```{r plotSmoothedDailyCases}

mkPlot(cases, ctys, col, "case", "Daily", lag=7, yaxt="log",
       day0=d.day0, D0=10, xlab="Days since 10th death")

```

#### Linear scale

```{r plotSmoothedDailyCasesLinear}

mkPlot(cases, ctys, col, "case", "Daily", lag=7, yaxt="lin",
       day0=d.day0, D0=10, xlab="Days since 10th death")

```


### Raw {.tabset .tabset-fade}

#### Log-scale

```{r plotDailyCases}

mkPlot(cases, ctys, col, "case", "Daily", lag=1, yaxt="log",
       day0=d.day0, D0=10, xlab="Days since 10th death")

```

#### Linear-scale

```{r plotDailyCasesLinear}

mkPlot(cases, ctys, col, "case", "Daily", lag=1, yaxt="lin",
       day0=d.day0, D0=10, xlab="Days since 10th death")

```




## Cumulative deaths {.tabset .tabset-fade}

### Log-scale

```{r plotSeveralAlign, fig.cap="Gray dashed lines are doubling times of 1, 2, 3, and 7 days (from steepest to shallowest)", results="hide"}

# we'll use this later for the time-scale for some other plots
D0 = 10
d.day0 = sapply(ctys, findTime, deaths=deaths, D0=D0)


mkPlot(deaths, ctys, col, "death", "Cumulative", lag=0, yaxt="log",
       day0=NULL, D0=10, xlab=NULL)

```

### Linear-scale

```{r plotSeveralAlignLinear, results="hide"}
mkPlot(deaths, ctys, col, "death", "Cumulative", lag=0, yaxt="lin",
       day0=NULL, D0=10, xlab=NULL)


```


## Cumulative cases {.tabset .tabset-fade}

### Log-scale

```{r plotSeveralAlignCase, fig.cap="Gray dashed lines are doubling times of 1, 2, 3, and 7 days (from steepest to shallowest)", results="hide"}

# we'll use this later for the time-scale for some other plots
C0 = 100
c.day0 = sapply(ctys, findTime, deaths=deaths, D0=C0)


mkPlot(cases, ctys, col, "case", "Cumulative", lag=0, yaxt="log",
       day0=NULL, D0=100, xlab=NULL)

```

### Linear-scale

```{r plotSeveralAlignCaseLinear, results="hide"}

mkPlot(cases, ctys, col, "case", "Cumulative", lag=0, yaxt="lin",
       day0=NULL, D0=100, xlab=NULL)

```





# Country-level naive case fatality rates {.tabset .tabset-fade}

By "naive" I mean: at any point in time, divide the total cumulative number of deaths
by the total cumulative number of confirmed cases.  This will be biased high
because the denominator is too small because not all cases are detected (due to
lack of testing), but on the other hand will be biased low because some current
active cases will result in deaths later.  It should eventually converge on the true CFR if testing
becomes widespread and the epidemic dies down.

## Log-scale

```{r plotCFR}
ndates = length(deaths$date)
adj = c(-0.1, 0.5)

cfr = deaths[,ctys] / cases[,ctys]
if ("China" %in% ctys) {
  xlim = c(0, max(deaths$doy) - sort(d.day0)[2] + 5)
} else {
  xlim = c(0, max(deaths$doy) - sort(d.day0)[1] + 5)
}
ylim = c(0, max(cfr, na.rm=TRUE))
ylim = c(5e-3, 0.2)

linlogplot(1, 1, typ='n', xlim=xlim, ylim=ylim, xlab="Days since 10th death", ylab="Naive CFR",
     main=paste("Naive CFR through", max(cases$date)), grid=TRUE)
for (icty in 1:length(ctys)) {
  lines(deaths$doy - d.day0[icty], cfr[,icty], col=col[icty])
  points(deaths$doy - d.day0[icty], cfr[, ctys[icty]], col=col[icty], cex=0.5, pch=16)
}

ilabel = pmin(ndates, floor(d.day0 + xlim[2]))
text(deaths$doy[ilabel] - d.day0,
     as.numeric(cfr[cbind(ilabel, 1:length(ctys))]),
     ctys, col=col, adj=adj, cex=0.6)


legend("topleft", legend=ctys, col=col, pch=16, cex=0.5, lty="solid")

```

## Linear-scale

```{r plotCFRLinear}
cfr = deaths[,ctys] / cases[,ctys]
ylim = c(0, max(cfr, na.rm=TRUE))
ylim = c(0, 0.2)

plot(1, 1, typ='n', xlim=xlim, ylim=ylim, xlab="Days since 10th death", ylab="Naive CFR",
     main=paste("Naive CFR through", max(cases$date)))
grid(NULL, col=gray(0.8), lty="solid", lwd=0.3)
for (icty in 1:length(ctys)) {
  lines(deaths$doy - d.day0[icty], cfr[,icty], col=col[icty])
  points(deaths$doy - d.day0[icty], cfr[, ctys[icty]], col=col[icty], cex=0.5, pch=16)
}

ilabel = pmin(ndates, floor(d.day0 + xlim[2]))
text(deaths$doy[ilabel] - d.day0,
     as.numeric(cfr[cbind(ilabel, 1:length(ctys))]),
     ctys, col=col, adj=adj, cex=0.6)


legend("topleft", legend=ctys, col=col, pch=16, cex=0.5, lty="solid")

```


# US state-level data

So far I'm using data collated by the NY Times.  It is pretty strangely structured,
but oh well:

```{r getUSData, echo=TRUE}
states = read.csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv",
                  stringsAsFactors=FALSE)
states$date = as.POSIXct(states$date)
states$doy = round(as.numeric(difftime(states$date, as.POSIXct("2019-12-31"), units="days")))

# Get this into a more sensible structure:
us.deaths = t(tapply(states$deaths,
                     list(factor(states$state, levels=sort(unique(states$state))),
                          factor(states$doy, levels=min(states$doy):max(states$doy))),
                     identity))
us.deaths = as.data.frame(us.deaths)
dates = sort(unique(states$date))
us.deaths$date = dates
doys = sort(unique(states$doy))
us.deaths$doy = doys

us.cases = t(tapply(states$cases,
                    list(factor(states$state, levels=sort(unique(states$state))),
                         factor(states$doy, levels=min(states$doy):max(states$doy))),
                    identity))
us.cases = as.data.frame(us.cases)
us.cases$date = dates
us.cases$doy = doys

```

# US state-level deaths and cases {.tabset .tabset-fade}

```{r setupStates}
sts = c("California", "Washington", "Oregon", "Colorado", "Arizona",
        "South Dakota", "North Dakota", "Wisconsin",
        "New York", "New Jersey", "Massachusetts",
        "Michigan", "Illinois", "Minnesota",
        "Florida", "Louisiana", "Georgia")

st.col = c(sapply(c(0.0, 0.1, 0.2, 0.3, 0.4)*2, function(f){whiten("orange1", f=f)}),
           sapply(c(0.0, 0.1, 0.2)*2, function(f){whiten("red3", f=f)}),
           sapply(c(0.0, 0.1, 0.2)*2, function(f){whiten("blue1", f=f)}),
           sapply(c(0.0, 0.1, 0.2)*2, function(f){whiten("purple1", f=f)}),
           sapply(c(0.0, 0.1, 0.2)*2, function(f){whiten("green1", f=f)}))

```

## Daily deaths {.tabset .tabset-fade}

### Smoothed {.tabset .tabset-fade}

#### Log-scale

```{r plotSmoothedDailyDeathsState}

mkPlot(us.deaths, sts, st.col, "death", "Daily", lag=7, yaxt="log",
       day0=d.day0, D0=1, xlab="Days since 10th death")

```

#### Linear scale

```{r plotSmoothedDailyDeathsLinearState}

mkPlot(us.deaths, sts, st.col, "death", "Daily", lag=7, yaxt="lin",
       day0=d.day0, D0=1, xlab="Days since 10th death")

```


### Raw {.tabset .tabset-fade}

#### Log-scale

```{r plotDailyDeathsState}

mkPlot(us.deaths, sts, st.col, "death", "Daily", lag=1, yaxt="log",
       day0=d.day0, D0=1, xlab="Days since 10th death")

```

#### Linear-scale

```{r plotDailyDeathsLinearState}

mkPlot(us.deaths, sts, st.col, "death", "Daily", lag=1, yaxt="lin",
       day0=d.day0, D0=1, xlab="Days since 10th death")


```


## Daily new cases {.tabset .tabset-fade}

### Smoothed {.tabset .tabset-fade}

#### Log-scale

```{r plotSmoothedDailyCasesState}

mkPlot(us.cases, sts, st.col, "case", "Daily", lag=7, yaxt="log",
       day0=d.day0, D0=10, xlab="Days since 10th death")

```

#### Linear scale

```{r plotSmoothedDailyCasesLinearState}

mkPlot(us.cases, sts, st.col, "case", "Daily", lag=7, yaxt="lin",
       day0=d.day0, D0=10, xlab="Days since 10th death")

```


### Raw {.tabset .tabset-fade}

#### Log-scale

```{r plotDailyCasesState}

mkPlot(us.cases, sts, st.col, "case", "Daily", lag=1, yaxt="log",
       day0=d.day0, D0=10, xlab="Days since 10th death")

```

#### Linear-scale

```{r plotDailyCasesLinearState}

mkPlot(us.cases, sts, st.col, "case", "Daily", lag=1, yaxt="lin",
       day0=d.day0, D0=10, xlab="Days since 10th death")

```



## Cumulative deaths {.tabset .tabset-fade}

### Log-scale

```{r plotSeveralAlignState, fig.cap="Gray dashed lines are doubling times of 1, 2, 3, and 7 days (from steepest to shallowest)", results="hide"}

# we'll use this later for the time-scale for some other plots
D0 = 10
d.day0 = sapply(sts, findTime, deaths=us.deaths, D0=D0)


mkPlot(us.deaths, sts, st.col, "death", "Cumulative", lag=0, yaxt="log",
       day0=NULL, D0=10, xlab=NULL)

```

### Linear-scale

```{r plotSeveralAlignLinearState, results="hide"}
mkPlot(us.deaths, sts, st.col, "death", "Cumulative", lag=0, yaxt="lin",
       day0=NULL, D0=10, xlab=NULL)


```

## Cumulative cases {.tabset .tabset-fade}

### Log-scale

```{r plotSeveralAlignCaseState, fig.cap="Gray dashed lines are doubling times of 1, 2, 3, and 7 days (from steepest to shallowest)", results="hide"}

mkPlot(us.cases, sts, st.col, "case", "Cumulative", lag=0, yaxt="log",
       day0=NULL, D0=100, xlab=NULL)

```

### Linear-scale

```{r plotSeveralAlignCaseLinearState, results="hide"}

mkPlot(us.cases, sts, st.col, "case", "Cumulative", lag=0, yaxt="lin",
       day0=NULL, D0=100, xlab=NULL)

```




# US state-level naive case fatality rates {.tabset .tabset-fade}

## Log-scale

```{r plotCFRState}
ndates = length(us.deaths$date)

us.cfr = us.deaths[,sts] / us.cases[,sts]
xlim = c(0, max(us.deaths$doy) - sort(d.day0)[1] + 5)
ylim = c(0, max(us.cfr, na.rm=TRUE))
ylim = c(5e-3, 0.15)

linlogplot(1, 1, typ='n', xlim=xlim, ylim=ylim, xlab="Days since 10th death", ylab="Naive CFR",
     main=paste("Naive CFR through", max(us.cases$date)), grid=TRUE)
for (icty in 1:length(sts)) {
  lines(us.deaths$doy - d.day0[icty], us.cfr[,icty], col=st.col[icty])
  points(us.deaths$doy - d.day0[icty], us.cfr[, sts[icty]], col=st.col[icty], cex=0.5, pch=16)
}

ilabel = pmin(ndates, floor(d.day0 + xlim[2]))
text(us.deaths$doy[ilabel] - d.day0,
     as.numeric(us.cfr[cbind(ilabel, 1:length(sts))]),
     sts, col=st.col, adj=adj, cex=0.6)


legend("topleft", legend=sts, col=st.col, pch=16, cex=0.5, lty="solid")

```

## Linear-scale

```{r plotCFRLinearState}
ylim = c(5e-3, 0.15)

plot(1, 1, typ='n', xlim=xlim, ylim=ylim, xlab="Days since 10th death", ylab="Naive CFR",
     main=paste("Naive CFR through", max(us.cases$date)))
grid(NULL, col=gray(0.8), lty="solid", lwd=0.3)
for (icty in 1:length(sts)) {
  lines(us.deaths$doy - d.day0[icty], us.cfr[,icty], col=st.col[icty])
  points(us.deaths$doy - d.day0[icty], us.cfr[, sts[icty]], col=st.col[icty], cex=0.5, pch=16)
}

ilabel = pmin(ndates, floor(d.day0 + xlim[2]))
text(us.deaths$doy[ilabel] - d.day0,
     as.numeric(us.cfr[cbind(ilabel, 1:length(sts))]),
     sts, col=st.col, adj=adj, cex=0.6)


legend("topleft", legend=sts, col=st.col, pch=16, cex=0.5, lty="solid")

```
